Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-06-02 cordillera-cycle.tex
Update discussion outline.
Corrections from Irina in intro and setup.

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera were more extensive than today. During the last glacial maximum, a
contiguous blanket of ice extended from Denali Peak in the north to the Puget
Sound in the south \aref. This ice mass is known as the former Cordilleran ice
sheet (Fig.~\ref{fig:locmap}).

For more than a century, exploration and geological investigation of the
Cordillera have led to a large collection of evidence of the former ice cover \aref. This
evidence consists of mapped boundaries of the former ice extent \aref, directions of
past sliding velocities \aref and locations of former melt-water streams \aref, as well as
the timing of glacier retreat identified from radiocarbon and cosmogenic dating \aref.

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last glacial
cycle \aref. However, former ice thickness and the ice sheet's contribution to the
last glacial maximum sea-level low-stand remain poorly constrained. Moreover,
our understanding of the glaciation history is mainly restricted to the
deglaciation phase and on a regional scale, while little is known about the ice
sheet evolution prior to the last glacial maximum extent \aref. Although
time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
patterns are available for the Laurentide and Eurasian ice sheets \aref, such is not
the case for the Cordilleran ice sheet, where complex arrangements of ice flow
directions emanating from multiple glaciation centres have been identified \aref.

@@

The simulations presented here were run using Parallel Ice Sheet Model (PISM,
development version~11b0a7f), an open-source, finite-difference, shallow ice
sheet model \aref. The model takes basal topography, sea level, geothermal heat flux
and climate forcing, as input fields and computes the evolution of ice extent
and thickness in time, its thermal and dynamic state, and
the associated lithospheric response. Our modelling domain encompasses the
entire area covered by the Cordilleran ice sheet at the last glacial maximum
(Fig.~\ref{fig:locmap}).

@@

\note{Do we need more details (equations) on this new sliding law?}

Sea level is lowered as a function of time according to the SPECMAP
reconstruction, and basal topography responds to ice load
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
half-space \citep{lingle-clark-1985,bueler-etal-2007}. It uses a lithosphere
density of 3300\,\unit{kg\,m^{-3}}, a flexural rigidity of $5 \times
10^{24}$\,\unit{N\,m} and a mantle viscosity of $1 \times
10^{21}$\,\unit{Pa\,s}. Due to the high mantle viscosity, there exists a time
lag between ice sheet growth and isostatic bedrock response.

@@

Atmospheric forcing of the model consists of a present-day monthly climatology
modified by time-dependent temperature offsets and a lapse-rate correction.
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t) + {\Delta}T_{LR}(t, x, y) \\
    P_m(t, x, y) &= P_{m0}(x, y)
\end{align}

The present-day climatology $\{T_m, P_m\}$ was computed from the near-surface air temperature and
precipitation rate fields of the North American Regional Reanalysis (NARR) over
the period 1979--2000 (Fig.~\ref{fig:atm}). Modern climate of the North
American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability.
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as outlined in a previous sensitivity
study.

@@

\subsection{Ice-sheet scale glaciation dynamics}

@@

@@

\subsection{Deglaciation patterns}

>>>>

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera were more extensive than today. At the last glacial maximum, a
contiguous blanket of ice extended from Denali Peak in the north to the Puget
Sound in the south \aref. This ice mass is known as the former Cordilleran ice
sheet (Fig.~\ref{fig:locmap}).
\note{Make sure these locations are marked on Fig.~\ref{fig:locmap}.}

For more than a century, exploration and geological investigation of the
Cordillera have led to a large collection of evidence of the former ice cover \aref. This
evidence consists of mapped boundaries of the former ice extent \aref, directions of
past sliding velocities {\aref} and locations of former melt-water streams \aref, as well as
the timing of glacier retreat identified from radiocarbon and cosmogenic dating \aref.

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last glacial
cycle \aref. However, former ice thickness and the ice sheet's contribution to the
last glacial maximum sea-level low-stand remain poorly constrained. Moreover,
our understanding of the glaciation history is mainly restricted to the
deglaciation phase and a regional scale, while little is known about the ice
sheet evolution prior to the last glacial maximum extent \aref. Although
time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
patterns are available for the Laurentide and Eurasian ice sheets \aref, such is not
the case for the Cordilleran ice sheet, where complex arrangements of ice flow
directions emanating from multiple glaciation centres have been identified \aref.

@@

The simulations presented here were run using Parallel Ice Sheet Model (PISM,
development version~11b0a7f), an open-source, finite-difference, shallow ice
sheet model \aref. The model takes basal topography, sea level, geothermal heat flux
and climate forcing, as input fields and computes the evolution of ice extent
and thickness in time, the thermal and dynamic state of the ice sheet, and
the associated lithospheric response. Our modelling domain encompasses the
entire area covered by the Cordilleran ice sheet at the last glacial maximum
(Fig.~\ref{fig:locmap}).

@@

Sea level is lowered as a function of time according to the SPECMAP
reconstruction, and basal topography responds to ice load
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
half-space \citep{lingle-clark-1985,bueler-etal-2007}.

@@

Atmospheric forcing of the model consists of a present-day monthly climatology
$\{T_m0, P_m0\}$, modified by temperature-offset time series ${\Delta}T_{TS}$
and a lapse-rate correction ${\Delta}T_{LR}$.
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t) + {\Delta}T_{LR}(t, x, y) \\
    P_m(t, x, y) &= P_{m0}(x, y)
\end{align}

The present-day climatology $\{T_m0, P_m0\}$ was computed from the near-surface air temperature and
precipitation rate fields of the North American Regional Reanalysis (NARR) over
the period 1979--2000 (Fig.~\ref{fig:atm}). Modern climate of the North
American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability.
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as outlined in a previous sensitivity
study.

@@

\subsection{Timing of the glacial maxima}

@@

\subsection{Glaciated areas through the last glacial cycle}

@@

\subsection{History of the last deglaciation}
